Quantifying Infants' Everyday Restrained Experiences in the Home Using Wearable Inertial Sensors

### 2025-11-09 manuscript_restrained_classification validation.qmd
fixed typo

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). Model evaluation, statistical analyses, and this reproducible manuscript were conducted in RStudio environment (R version 4.5.0; RStudio version 2025.5.0.496; RStudioTeam, 2024) using the R packages tidyverse [@tidyverse], flextable [@flextable], and papaja [@papaja], as well as the apa quarto template (Schneider, n.d.). All human annotation results and processing scripts are publicly shared on GitHub (XXX).

@@

The current work also demonstrates that wearable sensors have the potential to characterize other categories of infant movement that might interest researchers in a variety of domains. Compared with previous works that classify infants’ positions [@Airaksinen2020j; @Airaksinen2024z; @Franchak2021k; @Franchak2024e; @Yao2019q], restraint might be the most challenging one out of all categories, because the motion between classes can be very similar sometimes—such as sitting on the ground versus sitting on a caregiver’s lap. In contrast, other categories (e.g., positions) usually have distinctive motion features between classes—such as the legs’ moving direction in supine position versus upright position. Our results, however, showed that even a context-dependent category as restraint can be successfully classified by machine learning models using sensor data. Conceivably, this system might perform comparably well in classifying other context-independent categories, such as locomotion, steps, falling, etc. Future research needs to further develop the potential of wearable sensors. If proven feasible, the new models will substantially reduce the need for human labor and enlarge the data scale.

Beyond expanding the range of classifiable behaviors, future works should also broaden the scenarios in which wearable sensors are applied. To date, most applications of wearable sensors have been limited to laboratory or home environments. Infants’ outdoor experiences remain largely unexplored, whereas outdoor environments provide infants with rich and unique opportunities for learning and exploration. How much time infants spend restrained in stroller or car seat? How positions are they frequently in? How many steps they take? Validating sensor-based models in outdoor scenarios can open new avenues for investigation.

Moving forward, future works can also integrate different categories of model predictions to generate a multifaceted description of infant real-time experience. For example, by combing the model prediction of infant positions (i.e., supine, prone, sitting, upright, holding and carrying) with the current model prediction of overall restraint, researchers can easily know what positions infants were in while restrained—such as supine restrained, sitting restrained, or being held—without spending extra efforts in human coding and model training.

@@

The advances in methodology can facilitate a deeper understanding of infant full-day real-time restraint experiences. Restraint, as an external factor imposed by caregivers, plays a critical role in infant real-time experiences—influencing when and how often infants have the opportunities for movement, which portions of the environment they can perceive, and what objects and people they can interact with. Prior works provided valuable insights into parental attitudes of restraint and physical activity. For example, parental concerns over floor play, beliefs about physical activeness, and practices of co-play are associated with restraint frequency and infants’ physical activity level [@Hanisch2025j; @Hnatiuk2013t; @Prioreschi2017c; @Hesketh2015i; @Karasik2025h]. However, few studies have examined how each restraint decision is formed in real time and how infants’ sensorimotor experiences are affected in real time.

Having a full-day continuous profile of infant restraint is thus essential in identifying the factors influencing restraint decisions in real time. Restraint is believed to serve practical purposes such as safety, transportation, comforting, etc. [@Birken2015s], but these functions alone cannot account for all the variability in when caregivers choose to restrain an infant. To what extent are such decisions influenced by these intra-individual factors, such as daily routines, momentary tasks, or fluctuations in emotional states? To what extent are such decisions influenced by inter-individual factors, such as infant motor skill, attachment style, caregivers’ beliefs, or parental stress? And to what extent are the decisions influenced by socio-cultural factors, such as cultural norms [@Karasik2015i; @Karasik2018t], or socioeconomic status? The current system’s capability of locating each individual bout of restraint enables more fine-grained analyses of caregiver behavior and provides new insights into the dynamics of everyday caregiving practices.

Having a full-day continuous profile of infant restraint is also essential in understanding infant real-time sensorimotor experiences. An overall estimate of restraint time is not enough to examine the mechanisms between restraint practices and development outcomes. Take motor development as an example. @Karasik2023d found that Tajik infants fell behind US infants in achieving motor milestones despite the comparable overall restrained time between cultural groups, might due to the prolonged individual restrained bouts in Tajik infants [@Karasik2022a]. The current system allows researchers to characterize the temporal structure of restraint and examine the relationship between restraint and other experiences. A prior study [@Malachowski2023a] found device restraint is associated with reduced adult language input using ecological momentary assessment. However, a continuous, full-day recording of restraint captures the temporal distribution rather than snapshots. Questions can be answered such as: Does adult word count vary based on the duration of each restraint bout? How do specific language inputs (e.g., referential words) distribute throughout a day based on restraint states and restraint types? Additionally, real-time restraint can also be linked to infants’ self-initiated explorations. For example, how do infants’ own vocalizations, attention, locomotion, or object interactions differ across restrained and unrestrained? In short, continuous, full-day quantification of infant restraint offers a foundation for linking caregiver practices to infants’ moment-to-moment sensorimotor experiences, thereby deepening our understanding of how daily contexts accumulate shape developmental outcomes.

@@

To conclude, the current study provided a powerful tool for quantifying infants’ everyday restraint experiences in the home using wearable sensors and machine learning modeling. The model achieved high accuracy and strong agreement with human annotations, as well as convergent validity by showing an age-related decrease in infant restraint time. We explored the window length in the data aggregating procedure of modeling and proved that a longer window enhanced the model’s ability to capture the contextual heterogeneous nature of restraint. Together, the current study advanced the method for measuring infant full-day, real-time movement, and opened avenues for research on how external constraints from caregivers shape early development.

>>>>

Machine-learning process was conducted in Julia version 1.11.1 using the DecisionTree package (Sadeghi et al., 2022). Model evaluation, statistical analyses, and this reproducible manuscript were conducted in RStudio environment (R version 4.5.0; RStudio version 2025.5.0.496; RStudioTeam, 2024) using the R packages tidyverse [@tidyverse], flextable [@flextable], and papaja [@papaja], as well as the apa quarto template (Schneider, n.d.). All human annotation results and processing scripts are publicly shared on GitHub (https://github.com/Hanzhi-W/ms_restrained_classification-validation.git).

@@

The current work also demonstrates that wearable sensors have the potential to characterize other categories of infant movement that might interest researchers in a variety of domains. Compared with previous works that classify infants’ positions [@Airaksinen2020j; @Airaksinen2024z; @Franchak2021k; @Franchak2024e; @Yao2019q], restraint might be the most challenging one out of all categories, because the motion between classes can be very similar sometimes—such as sitting on the ground versus sitting on a caregiver’s lap. In contrast, positions usually have distinctive motion features between classes—such as the legs’ moving direction in supine position versus upright position. Our results, however, showed that even a context-dependent category as restraint can be successfully classified by machine learning models using sensor data. Conceivably, this system might perform comparably well or better in classifying other context-independent categories, such as locomotion, steps, falling, etc. Future research needs to further develop the potential of wearable sensors. If proven feasible, the new models will substantially reduce the need for human labor in coding and enlarge the data scale.

Beyond expanding the range of classifiable behaviors, future works should also broaden the scenarios in which wearable sensors are applied. To date, most applications of wearable sensors have been limited to laboratory or in-home environments. Infants’ outdoor experiences remain largely underexplored, whereas outdoor environments provide infants with rich and unique opportunities for learning and exploration [@Schneider2025d; @Herrington1998o; @Fjortoft2000o]. How much time infants spend restrained in stroller or car seat? Which positions are they frequently in? How many steps they take? Validating sensor-based models in outdoor scenarios can open new avenues for investigation.

Moving forward, future works can also integrate different categories of model predictions to generate a multifaceted description of infant real-time experiences. For example, by combing the model prediction of infant positions (i.e., supine, prone, sitting, upright, holding and carrying) with the current model prediction of overall restraint, researchers can easily know what positions infants were in while restrained—such as supine restrained, sitting restrained, or being held—without spending extra efforts in human coding and model training.

@@

The advances in methodology can facilitate a deeper understanding of infant full-day real-time restraint experiences. Restraint, as an external factor imposed by caregivers, plays a critical role in infant real-time experiences—influencing when and how often infants have the opportunities for movement, which portions of the environment they can perceive, and what objects and people they can interact with. Prior works provided valuable insights into parental attitudes of restraint and physical activity. For example, parental concerns over floor play, beliefs about physical activeness, and practices of co-play are found to be associated with restraint frequency and infants’ physical activity level [@Hanisch2025j; @Hnatiuk2013t; @Prioreschi2017c; @Hesketh2015i; @Karasik2025h]. However, few studies have examined how each restraint decision is formed in real time, and how infants’ sensorimotor experiences are affected in real time.

Having a full-day continuous profile of infant restraint is thus essential in identifying the factors influencing restraint decisions in real time. Restraint is believed to serve practical purposes such as safety, transportation, comforting, etc. [@Birken2015s], but these functions alone cannot account for all the variability in when caregivers choose to restrain an infant. To what extent are such decisions influenced by these intra-individual factors, such as daily routines, or momentary tasks? To what extent are such decisions influenced by inter-individual factors, such as infant motor skill, attachment style, caregivers’ beliefs, or parental stress? And to what extent are the decisions influenced by socio-cultural factors, such as cultural norms [@Karasik2015i; @Karasik2018t], or socioeconomic status? The current system’s capability of locating each individual bout of restraint enables more fine-grained analyses of caregiver behaviors and provides new insights into the dynamics of everyday caregiving practices.

Having a full-day continuous profile of infant restraint is also essential in understanding infant real-time sensorimotor experiences. An overall estimate of restraint time is not enough to examine the mechanisms between restraint practices and development outcomes. Take motor development as an example. @Karasik2023d found that Tajik infants fell behind US infants in achieving motor milestones despite the comparable overall restrained time between cultural groups. It is likely due to the prolonged individual restrained bouts in Tajik infants [@Karasik2022a]. The current system allows researchers to characterize the temporal structure of restraint and examine the relationship between restraint and other experiences, such as language, object interaction, and social interaction. A prior study [@Malachowski2023a] found device restraint is associated with reduced adult language input using ecological momentary assessment. However, a continuous, full-day recording of restraint can capture the temporal distribution rather than snapshots. Questions can be answered such as: Does adult word count vary based on the duration of each restraint bout? How do specific language inputs (e.g., referential words) distribute throughout a day based on restraint states and restraint types? Additionally, real-time restraint can also be linked to infants’ self-initiated explorations. For example, how do infants’ own vocalizations, attention, locomotion, or object interactions differ across restrained and unrestrained? In short, continuous, full-day quantification of infant restraint offers a foundation for linking caregiver practices to infants’ moment-to-moment sensorimotor experiences, thereby shaping infants' developmental outcomes.

@@

To conclude, the current study provided a powerful tool for quantifying infants’ everyday restraint experiences in the home using wearable sensors and machine learning modeling. The model achieved high accuracy and strong agreement with human annotations, as well as convergent validity by showing an age-related decrease in infant restraint time. We explored the window length in the data aggregating procedure of modeling and proved that a longer window enhanced the model’s balanced performance. Together, the current study advanced the method for measuring infant full-day, real-time movement, and opened avenues for research on how restraint imposed by caregivers shape early development.
